---
title: "Custom Databases: Classification & Analyses"
author: "Bastiaan"
date: "11 January 2021"
output: html_document
---

Introduction
------------

The following R markdown file performs the steps needed to search out
and visualize overlap / gaps between obtained public sequence data
and its reference provided by the Dutch Species Register (NSR).

The following libraries need to be loaded:

```{r libraries, include=FALSE, message=FALSE, warning=FALSE}
library(rmarkdown)
library(taxizedb)
library(myTAI)
library(tidyr)
library(shiny)
library(DT)
library(plyr)
library(dplyr)
library(d3Tree)
library(billboarder)
```

In addition, here We define the global variables that we will reuse throughout the code:

```{r globals, include=FALSE}
# This should define the root folder of the local copy of the git repository at: https://github.com/naturalis/Custom-databases-DNA-sequences, which is
# automatically defined correctly if we run the present code from within a
# local clone of the repo and have set the working directory to the script
# source (in RStudio: Session > Set working directory > To source file location)
REPO_HOME <- paste(dirname(getwd()), sep = "")
```

### Load data

We now load the following data sets:

1. Reference data set (NSR),
2. Obtained Public Sequence Data (BOLD),
3. List of Marker Codes

Both the Reference Data set (1) and the Obtained Public Sequence Data (2) are output from
our [script](../script/custom_databases.py)

```{r datasets, include=FALSE}
# Load species from the NSR file
species.file <- sprintf('%s/results/species.csv', REPO_HOME)
species <- read.csv(species.file, header=TRUE, sep=",")

# Load public sequence data
# Set missing values to na
bold.file <- sprintf('%s/data/FASTA_files/match.fasta', REPO_HOME)
bold <- read.csv(bold.file, header=TRUE, sep="\t", row.names=NULL,
                 na.strings=c(""," ","NA"))

# Load marker data set
markers.file <- sprintf('%s/results/markers.csv', REPO_HOME)
markers <- read.csv(markers.file, header=TRUE, sep=",")
```

### Classification

Ideally a full taxonomic hierarchy of species would be used as reference data set. As
only the scientific names have been given, classification of these species will be
extracted from a database of choice (ncbi, itis, gbif, col, wfo).

```{r classification, include=FALSE}
# Download taxonomic database
# data source: NCBI
db_download_ncbi()

# Load species names from the NSR data set
taxidIn <- species[, "species_name"]

# Convert species names to taxon IDs
taxidOut <- taxizedb::name2taxid(taxidIn, db="ncbi", out_type="summary")

# Isolate taxon IDs from output, save as vector
treeIn <- as.vector(taxidOut$id)

# Retrieve taxonomic hierarchy for each taxon ID
treeOut <- taxizedb::classification(treeIn, db="ncbi", row=1, verbose=FALSE)

# Parse out the taxonomy levels that you require
taxdata = data.frame()
for(x in 1:length(treeOut)){
    tryCatch({
      c.phylum=filter(treeOut[[x]], rank =="phylum")$name
      c.class=filter(treeOut[[x]], rank =="class")$name
      c.order=filter(treeOut[[x]], rank =="order")$name
      c.family=filter(treeOut[[x]], rank =="family")$name
      c.genus=filter(treeOut[[x]], rank =="genus")$name
      c.species=filter(treeOut[[x]], rank =="species")$name

      row <- data.frame(cbind(phylum=c.phylum,class=c.class,order=c.order,family=c.family,
                              genus=c.genus,species=c.species))
      taxdata <- bind_rows(taxdata, row)
    }, error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
}
write.csv(taxdata, file=sprintf('%s/results/taxdata-full.csv', REPO_HOME),
          row.names=FALSE)

taxdataUnique <- unique(taxdata)

# Clean-up
rm(c.phylum,c.class,c.order,c.family,c.genus,c.species,x,row)
```

### Analyses

Now we will consolidate the three data sets, creating a more streamlined data structure
which allows us to track information between each source.

```{r combine, include=FALSE}
# Joining the initial loaded data sets on common data intersections
species_markers <- right_join(species, bold[c("species_name","markercode")], by=c("species_name")) %>%
  group_by_all() %>% summarise(.groups="keep", counts = n()) %>%
  right_join(., markers, by=c("markercode" = "marker_name"))

# Formatting; Remove unwanted columns, add unique prefix, re-sort
species_markers <- within(species_markers, {
  species_name <- NULL
  markercode <- NULL
  sm_id <- seq.int(nrow(species_markers))
}) %>% .[, c(4,1,3,2)] %>% .[complete.cases(.[ , 2]),]

# Write data as csv to output folder
write.csv(species_markers, file=sprintf('%s/results/species_markers.csv', REPO_HOME),
          row.names=FALSE)

# Save workspace image
save.image(sprintf('%s/script/workspace.RData', REPO_HOME))
```
