---
title: "Custom Databases Analyses"
author: "Bastiaan"
date: "11 January 2021"
output: html_document
---

Introduction
------------

The following R markdown file performs the steps needed to search out and visualize overlap / gaps between obtained public sequence data (using the custom_databases.py script) and its reference by the Dutch Species Register.

The following libraries and scripts need to be loaded:

```{r libraries, include=FALSE}
library(rmarkdown)
library(flexdashboard)
library(crosstalk)
library(collapsibleTree)
library(reshape2)
library(shiny)
library(stringr)
library(DT)
library(plyr)
library(dplyr)
library(d3Tree)
library(rhandsontable)
```


In addition, We define the global variables that allow us to locate our data:

```{r globals, include=FALSE}
# This should define the root folder of the local copy of the git repository at: https://github.com/naturalis/Custom-databases-DNA-sequences, which is
# automatically defined correctly if we run the present code from within a
# local clone of the repo and have set the working directory to the script
# source (in RStudio: Session > Set working directory > To source file location)

REPO_HOME <- paste(dirname(getwd()), sep = "")
```


### Public Sequence Data (BOLD)

We now load the list of taxa for the matching species.

```{r Data, include=FALSE}
bold.file1 <- sprintf('%s/data/FASTA_files/match.fasta', REPO_HOME)


# Taxa to table
taxa <- read.csv(bold.file1, sep = "\t", row.names=NULL)[c("phylum_name","class_name","order_name","family_name","genus_name","species_name")]

# Dataframe of unique taxa count
taxa_unique <- count(taxa, phylum_name,class_name,order_name,family_name,genus_name,species_name) %>% ungroup()
```


### Chart A

```{r}
# Compute and draw a collapsible tree
collapsibleTreeSummary(
  taxa_unique,
  hierarchy = c("phylum_name", "class_name", "order_name", "family_name", "genus_name", "species_name"),
  height = 1000,
  width = 1500,
  zoomable = TRUE,
  attribute = "n",
  maxPercent = 50
)

datatable(taxa_unique, options = list(pageLength = 5))
```