---
title: "Custom Databases: Classification & Analyses"
author: "Bastiaan"
date: "21 January 2021"
output: html_document
runtime: shiny
---

Introduction
------------

The following R markdown file performs the steps needed to:

1. Create a more streamlined data structure, which allows us to track all information
of the species of interest between the various datasets.
2. Search out and visualize the coverage between the obtained public sequence data and
its reference from the Dutch Species Register (NSR).

The following libraries need to be loaded:

```{r libraries, include=FALSE, message=FALSE, warning=FALSE}
library(rmarkdown)
library(taxizedb)
library(myTAI)
library(tidyr)
library(shiny)
library(DT)
library(plyr)
library(dplyr)
library(d3Tree)
library(billboarder)
```

In addition, here We define the global variables that we will reuse throughout
the code:

```{r globals, include=FALSE}
# This should define the root folder of the local copy of the git repository at:
# https://github.com/naturalis/Custom-databases-DNA-sequences, which is
# automatically defined correctly if we run the present code from within a
# local clone of the repo and have set the working directory to the script
# source (in RStudio: Session > Set working directory > To source file location)
REPO_HOME <- paste(dirname(getwd()), sep = "")
```

### Load data

We now load the following data sets:

1. Reference data set (NSR),
2. Obtained Public Sequence Data (BOLD),
3. List of Marker Codes,
4. List of Public Databases

Both the Reference Data set (1) and the Obtained Public Sequence Data (2) are
output from the [script](../script/custom_databases.py)

```{r datasets, include=FALSE}
# Load species from the NSR file
species.file <- sprintf('%s/results/species.csv', REPO_HOME)
species <- read.csv(species.file, header=TRUE, sep=",")

# Load public sequence data
# Set missing values to na
bold.file <- sprintf('%s/data/FASTA_files/match.fasta', REPO_HOME)
bold <- read.csv(bold.file, header=TRUE, sep="\t", row.names=NULL,
                 na.strings=c(""," ","NA"))

# Load marker data set
markers.file <- sprintf('%s/results/markers.csv', REPO_HOME)
markers <- read.csv(markers.file, header=TRUE, sep=",")

# Load database data set
databases.file <- sprintf('%s/results/databases.csv', REPO_HOME)
databases <- read.csv(databases.file, header=TRUE, sep=",")
```

### Classification

Ideally a full taxonomic hierarchy of species would be used as reference dataset.
As only the scientific names have been given, classification of these species
will be extracted from a database of choice (ncbi, itis, gbif, col, wfo).

```{r classification, include=FALSE}
# Download taxonomic database
# data source: NCBI
db_download_ncbi()

# Load species names from the NSR data set
taxidIn <- species[, "species_name"]

# Convert species names to taxon IDs
taxidOut <- taxizedb::name2taxid(taxidIn, db="ncbi", out_type="summary")

# Isolate taxon IDs from output, save as vector
treeIn <- as.vector(taxidOut$id)

# Retrieve taxonomic hierarchy for each taxon ID
treeOut <- taxizedb::classification(treeIn, db="ncbi", row=1, verbose=FALSE)

# Parse out the taxonomy levels that you require
taxdata = data.frame()
for(x in 1:length(treeOut)){
  tryCatch({
    c.phylum=filter(treeOut[[x]], rank =="phylum")$name
    c.class=filter(treeOut[[x]], rank =="class")$name
    c.order=filter(treeOut[[x]], rank =="order")$name
    c.family=filter(treeOut[[x]], rank =="family")$name
    c.genus=filter(treeOut[[x]], rank =="genus")$name
    c.species=filter(treeOut[[x]], rank =="species")$name
    
    row <- data.frame(cbind(phylum=c.phylum,class=c.class,order=c.order,
                            family=c.family,genus=c.genus,species=c.species))
    taxdata <- bind_rows(taxdata, row)
  }, error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
}

# Write unique hierarchies to file
taxdata <- unique(taxdata)
write.csv(taxdata, file=sprintf('%s/results/taxdata-full.csv', REPO_HOME),
          row.names=FALSE)

# Clean-up
rm(c.phylum,c.class,c.order,c.family,c.genus,c.species,x,row)
```

### Create data structure

Now we will consolidate the three data sets, creating a more streamlined
data structure which allows us to track information between each source.

```{r combine, include=FALSE}
# Joining the initial loaded data sets on common data intersections
species_markers <- right_join(species, bold[c("species_name","markercode")],
                              by=c("species_name")) %>%
  group_by_all() %>% summarise(.groups="keep", counts = n()) %>%
  right_join(., markers, by=c("markercode" = "marker_name"))

# Formatting
species_markers <- within(species_markers, {
  #Remove obsolete columns
  species_name <- NULL
  markercode <- NULL
  
  # Add unique prefix and database ID
  sm_id <- seq.int(nrow(species_markers))
  database_id <- 1
}) %>% .[, c(5,1,4,3,2)] %>% .[complete.cases(.[ , 2]),] # Rearrange and remove NAs

# Write data as csv to output folder
write.csv(species_markers, file=sprintf('%s/results/species_markers.csv',
                                        REPO_HOME),row.names=FALSE)

# Save workspace image
save.image(sprintf('%s/script/workspace.RData', REPO_HOME))
```

### Load Tree

To perform an analyses on our species we now load the tree for our reference data set. This
is their complete tree populated with our collected data, where we seek out overlap / gaps
between the two. We will reconcile these data sets later on.

```{r tree-data, include=FALSE}
# Dataframe: Get backbone, marker names, and count for each species
vis.tree <- inner_join(species, taxdata, by=c("species_name" = "species")) %>%
  left_join(., species_markers[c("species_id","database_id", "marker_id","counts")],
            by=c("species_id")) %>%
  left_join(., markers, by=c("marker_id")) %>%
  left_join(., databases, by=c("database_id"))

# Formatting
vis.tree <- within(vis.tree, {
  
  # Set no matching species to count 0
  f <- is.na(counts) == TRUE
  counts[f] <- 0
  
  # Remove obsolete rows
  species_id <- NULL
  marker_id <- NULL
  database_id <- NULL
  f <- NULL
  
}) %>% .[, c(2,3,4,5,6,1,7,8,9)] # Rearrange
```

### Perform visualisation

Compute and draw reactive shiny filters through collapsible d3js tree

```{r tree-visualisation}
m=vis.tree%>%data.frame%>%mutate(NEWCOL=NA)%>%distinct

ui <- fluidPage(
  fluidRow(
    column(6,
           uiOutput("Hierarchy"),
           tableOutput("clickView"),
           d3treeOutput(outputId="d3", height="800px")
    ),
    column(6,
           "Reactive Table",
           div(DT::dataTableOutput(outputId="table"),
               style = "font-size:80%")
           
    )
  ),
  
  hr(),
  
  fluidRow(
    column(8,billboarderOutput(outputId="billboarder")
    ),
    column(4,billboarderOutput(outputId="markerPie"))
  )
)

server <- function(input, output, session) {
  
  #SearchTree----
  
  output$Hierarchy <- renderUI({
    Hierarchy=names(m)
    Hierarchy=head(Hierarchy,-1)
    selectizeInput("Hierarchy","Tree Hierarchy",
                   choices = Hierarchy,multiple=T,selected = Hierarchy,
                   options=list(plugins=list('drag_drop','remove_button')))
  })
  
  network <- reactiveValues()
  
  observeEvent(input$d3_update,{
    network$nodes <- unlist(input$d3_update$.nodesData)
    activeNode<-input$d3_update$.activeNode
    if(!is.null(activeNode)) network$click <- jsonlite::fromJSON(activeNode)
  })
  
  observeEvent(network$click,{
    output$clickView<-renderTable({
      as.data.frame(network$click)
    },caption='Last Clicked Node',caption.placement='top')
  })
  
  
  TreeStruct=eventReactive(network$nodes,{
    df=m
    if(is.null(network$nodes)){
      df=m
    }else{
      
      x.filter=tree.filter(network$nodes,m)
      df=ddply(x.filter,.(ID),function(a.x){m%>%filter_(.dots = list(a.x$FILTER))%>%distinct})
    }
    df
  })
  
  observeEvent(input$Hierarchy,{
    output$d3 <- renderD3tree({
      if(is.null(input$Hierarchy)){
        p=m
      }else{
        p=m%>%select(one_of(c(input$Hierarchy,"NEWCOL")))%>%unique
      }
      
      d3tree(data = list(root = df2tree(struct = p,rootname = 'm'), layout = 'collapse'),
             activeReturn = c("phylum", "class", "order", "family", "genus", "species_name",
                              "count", "marker_name", "database_name" ))
    })
  })
  
  observeEvent(network$nodes,{
    output$results <- renderPrint({
      str.out=''
      if(!is.null(network$nodes)) str.out=tree.filter(network$nodes,m)
      return(str.out)
    })    
  })
  
  output$table <- DT::renderDataTable(expr = {
    TreeStruct()%>%select(-NEWCOL)
  },
  options=list(scrollX=T,title="Reactive Table")
  )
  
  output$billboarder <- renderBillboarder(expr = {
    dfActive <- TreeStruct()%>%select(-NEWCOL)
    nsrTotal <- unlist(lapply(dfActive[1:6], function(x) length(unique(x))))
    
    boldTotal <- subset(dfActive,database_name=="BOLD")
    boldTotal <- unlist(lapply(boldTotal[1:6], function(x) length(unique(x))))
    boldCov <- as.vector(boldTotal / nsrTotal * 100)
    
    vis.cov <- data.frame(
      taxa = paste(names(m)[0:6]),
      nsr_total = nsrTotal,
      bold_total = boldCov
    )
    
    billboarder() %>%
      bb_barchart(
        data = vis.cov[, c("taxa", "bold_total")]
      ) %>%
      bb_data(
        names = list(bold_total = "bold")
      ) %>%
      bb_y_grid(show = TRUE) %>%
      bb_y_axis(tick = list(format = suffix("%")),
                label = list(text = "Percentage of taxa covered", position = "outer-top")) %>%
      bb_legend(position = "inset", inset = list(anchor = "top-right")) %>%
      bb_labs(title = "Coverage of taxa")
  })
  
  output$markerPie <- renderBillboarder(expr = {
    dfActive <- TreeStruct()%>%select(-NEWCOL)
    vis.marker <- setNames(aggregate(dfActive$counts, list(dfActive$marker_name),
                                     FUN=sum), c("marker_name", "counts"))
    
    billboarder() %>%
      bb_piechart(data = vis.marker) %>%
      bb_labs(title = "Share of markers")
  })
  
}

shinyApp(ui, server)
```
