---
title: "Custom Databases: Classification & Analyses"
author: "Bastiaan"
date: "16 February 2021"
output: html_document
runtime: shiny
---

Introduction
------------

The following R markdown file performs the steps needed to:

1. Create a data structure capable of providing highly visual representations and at the
same time maintaining integrity of each data set’s origin.
2. Search out and visualize the overlap / gaps between the obtained Public Sequence Data
(BOLD) and a reference provided by the Naturalis Biodiversity center, in correlation to
the NSR.
3. Assess the quality and reliability of public sequence data.

The following libraries need to be loaded:

```{r libraries, include=FALSE, message=FALSE, warning=FALSE}
library(rmarkdown)
library(taxizedb)
library(myTAI)
library(tidyr)
library(shiny)
library(DT)
library(plyr)
library(dplyr)
library(d3Tree)
library(billboarder)
library(nbaR)
```

In addition, here We define the global variables that we will reuse throughout
the code:

```{r globals, include=FALSE}
# This should define the root folder of the local copy of the git repository at:
# https://github.com/naturalis/Custom-databases-DNA-sequences, which is
# automatically defined correctly if we run the present code from within a
# local clone of the repo and have set the working directory to the script
# source (in RStudio: Session > Set working directory > To source file location)
REPO_HOME <- paste(dirname(getwd()), sep = "")
```

### Load data

We now load the following data sets:

1. Reference Data Set (NSR),
2. Obtained Public Sequence Data (BOLD),
3. List of Marker Codes,
4. List of Public Databases

Both the Reference Data Set (1) and the Obtained Public Sequence Data (2) are
output from the [script](../script/custom_databases.py)

```{r datasets, include=FALSE}
# Load species and synonyms from the NSR files
nsrSpecies.file <- sprintf('%s/results/nsr_species.csv', REPO_HOME)
nsrSpecies <- read.csv(nsrSpecies.file)

nsrSynonyms.file <- sprintf('%s/data/NSR_exports/synonyms_extract.csv', REPO_HOME)
nsrSynonyms <- read.csv(nsrSynonyms.file)

# Load public sequence data
# Set missing values to na
bold.file <- sprintf('%s/data/FASTA_files/match.tsv', REPO_HOME)
bold <- read.csv(bold.file, header=TRUE, sep="\t", row.names=NULL,
                 na.strings=c(""," ","NA"))

# Load marker data set
markers.file <- sprintf('%s/results/markers.csv', REPO_HOME)
markers <- read.csv(markers.file, header=TRUE, sep=",")

# Load database data set
databases.file <- sprintf('%s/results/databases.csv', REPO_HOME)
databases <- read.csv(databases.file, header=TRUE, sep=",")
```

### NSR Backbone

Ideally, a full taxonomic hierarchy of species would be used as a reference data set. As
the NSR export only held the scientific names of all species of interest, the Netherlands
Biodiversity API will be queried to obtain their higher taxa.

```{r nsrBackbone, include=FALSE}
## Taxonomy to retrieve the hierarchy from
nsrIn <- tolower(nsrSpecies[, "species_name"])

## Query NSR taxon records (duration: ~40m)
queryParams <- list("sourceSystem.code"="NSR")
nsrSpecimens <- data.frame()
for(x in 1:length(nsrIn)){
  queryParams[["acceptedName.scientificNameGroup"]] <- nsrIn[x]
  nsrSpecimens <- bind_rows(nsrSpecimens, taxon_query(queryParams))
}

## Access required taxonomic fields
nsrBackbone = data.frame()
for(x in 1:nrow(nsrSpecimens)){
  tryCatch({
    c.kingdom=nsrSpecimens$defaultClassification$kingdom[x]
    c.phylum=nsrSpecimens$defaultClassification$phylum[x]
    c.class=nsrSpecimens$defaultClassification$className[x]
    c.order=nsrSpecimens$defaultClassification$order[x]
    c.family=nsrSpecimens$defaultClassification$family[x]
    c.genus=nsrSpecimens$defaultClassification$genus[x]
    c.species=nsrSpecimens$acceptedName$specificEpithet[x]
    c.authority=nsrSpecimens$acceptedName$authorshipVerbatim[x]
    
    row <- data.frame(cbind(kingdom=c.kingdom,phylum=c.phylum,class=c.class,order=c.order,
                            family=c.family,genus=c.genus,species=paste(c.genus,c.species),
                            identification_reference=c.authority))
    nsrBackbone <- bind_rows(nsrBackbone, row)
  }, error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
}

## Remove parenthesis
nsrBackbone$identification_reference = gsub("[()]", "", nsrBackbone$identification_reference)

## Clean up query variables
rm(x, row, queryParams, list = ls()[grep("^c.", ls())])
```

### Naturalis Coverage

To determine the overlap / gaps between the obtained public sequence data and our reference
at the Naturalis Biodiversity Center, in relation to the NSR, we query the Naturalis specimen
records (using the Netherlands Biodiversity API) to obtain their coverage of species.

```{r natCoverage, include=FALSE}
## Taxonomy to retrieve the hierarchy from
natIn <- unique(gsub("([A-Za-z]+).*", "\\1", nsrSpecies$species_name))

## Query Naturalis specimen records (duration: ~20m)
queryParams <- list()
natSpecimens <- data.frame()
for(x in 1:length(natIn)){
  queryParams[["identifications.scientificName.genusOrMonomial"]] <- natIn[x]
  natSpecimens <- bind_rows(natSpecimens, specimen_query(queryParams))
}

## Access required taxonomic fields
natSpecies = data.frame()
for(x in 1:nrow(natSpecimens)){
  tryCatch({
    c.genus=natSpecimens$identifications[[x]]$defaultClassification$genus
    c.species=natSpecimens$identifications[[x]]$defaultClassification$specificEpithet
    c.count=natSpecimens$numberOfSpecimen[[x]]
    
    row <- data.frame(cbind(species=paste(c.genus,c.species),counts=c.count))
    natSpecies <- bind_rows(natSpecies, row)
  }, error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
}

## Remove NAs, sum unique species record counts, add DB identifier
natSpecies <- natSpecies[complete.cases(natSpecies), ]
natSpecies$counts <- as.numeric(as.character(natSpecies$count))
natSpecies <- aggregate(counts ~ species, data=natSpecies, FUN=sum)
natSpecies$database_name <- "NATURALIS"

## Adopt accepted name if observation of synonym
nsrSynonyms$species <- gsub("([A-Za-z]+.[a-z]+).*", "\\1", nsrSynonyms$Synonym)
nsrSynonyms$Taxon <- gsub("([A-Za-z]+.[a-z]+).*", "\\1", nsrSynonyms$Taxon)

natSynonyms <- left_join(natSpecies, nsrSynonyms[2:3],by=c('species'))
natSynonyms$Taxon <- coalesce(natSynonyms$Taxon, natSynonyms$species)
natSynonyms$species <- NULL
colnames(natSynonyms)[3] <- "species"
natSynonyms <- natSynonyms[,c(3,1,2)]

## Match species against the NSR
natCoverage <- left_join(nsrSpecies[c(2:3)], natSynonyms, by=c("species_name"="species"))

## Clean up query variables
rm(x, row, queryParams, list = ls()[grep("^c.", ls())])
```

### Additional taxonmy

The ability to divert to another taxonomy could prove useful in certain situations. 
Classification of species will hence be extrated from a database of choice for later use.

```{r altTaxdata, include=FALSE}
## Download taxonomic database (options: ncbi, itis, gbif, col, wfo)
## Used data source: NCBI
db_download_ncbi()

## Load species names from the NSR data set
taxidIn <- nsrSpecies[, "species_name"]

## Match species names to NCBI taxon IDs
taxidOut <- taxizedb::name2taxid(taxidIn, db="ncbi", out_type="summary")

## Isolate taxon IDs from output, save as vector
treeIn <- as.vector(taxidOut$id)

## Retrieve taxonomic hierarchy for each taxon ID
treeOut <- taxizedb::classification(treeIn, db="ncbi", row=1, verbose=FALSE)

## Parse out the taxonomy levels/factors that you require
altTaxdata = data.frame()
for(x in 1:length(treeOut)){
  tryCatch({
    c.phylum=filter(treeOut[[x]], rank =="phylum")$name
    c.class=filter(treeOut[[x]], rank =="class")$name
    c.order=filter(treeOut[[x]], rank =="order")$name
    c.family=filter(treeOut[[x]], rank =="family")$name
    c.genus=filter(treeOut[[x]], rank =="genus")$name
    c.species=filter(treeOut[[x]], rank =="species")$name
    c.taxID=filter(treeOut[[x]], rank =="species")$id
    
    row <- data.frame(cbind(phylum=c.phylum,class=c.class,order=c.order,
                            family=c.family,genus=c.genus,node_name=c.species,tax_id=c.taxID))
    altTaxdata <- bind_rows(altTaxdata, row)
  }, error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
}

## Match hierarchies to NSR species names by taxon IDs
## Accounting for use of synonym names within NCBI
altTaxdata <- unique(altTaxdata)
altTaxdata <- left_join(taxidOut, altTaxdata, by=c("id" = "tax_id")) %>%
  left_join(., nsrSpecies, by=c("name" = "species_name")) %>%
  .[, c(2,9,3,4,5,6,7,1,8)] # Reaarange columns

## Adjust Header
names(altTaxdata) <- c("tax_id", "species_id", "phylum", "class", "order", "family",
                         "genus", "species", "node_name")

## Write taxonomic data to file
write.csv(altTaxdata, file=sprintf('%s/results/taxdata.csv', REPO_HOME),
          row.names=FALSE)

## Clean-up
rm(x, row, list = ls()[grep("^c.", ls())])
```

### Create data structure

Now we will consolidate the various data sets, creating a more streamlined
data structure which allows us to track information between each source.

```{r species_markers, include=FALSE}
# Joining the initial loaded data sets (NSR Species & BOLD) on common data intersections
# Subsequently, change out markercode for marker names
species_markers <- right_join(nsrSpecies, bold[c("species_name","markercode","sequenceID",
                                                 "identification_reference")],
                              by=c("species_name", "identification_reference")) %>%
  right_join(., markers, by=c("markercode" = "marker_name"))

# Formatting
species_markers <- within(species_markers, {
  #Remove obsolete columns
  species_name <- NULL
  markercode <- NULL
  identification_reference <- NULL
  
  # Add unique prefix and database ID (in case of BOLD: "1")
  sm_id <- seq.int(nrow(species_markers))
  database_id <- 1
}) %>% .[, c(5,1,4,3,2)] %>% .[complete.cases(.[ , 2]),] # Rearrange and remove NAs

# Write data as csv to output folder
write.csv(species_markers, file=sprintf('%s/results/species_markers.csv',
                                        REPO_HOME),row.names=FALSE)
```

### Load Tree

To perform an analyses on our species we now load the backbone of the NSR. This is their
complete tree, to be populated with our collected data where we seek out overlap / gaps
between Naturalis and our obtained Public Sequence Data .

```{r finalTree, include=FALSE}
## Get backbone of species, and add coverage of BOLD recrods
boldTree <- left_join(nsrSpecies, nsrBackbone[c(1:7)], by=c("species_name"="species")) %>%
  left_join(., species_markers[c("species_id","database_id", "marker_id")],
            by=c("species_id")) %>%
  left_join(., markers, by=c("marker_id")) %>%
  left_join(., databases, by=c("database_id"))

## Formatting: Remove obsolete rows
boldTree <- within(boldTree, {
  species_id <- NULL
  identification_reference <- NULL
  marker_id <- NULL
  database_id <- NULL
}) %>% .[, c(2,3,4,5,6,7,1,8,9)] # Rearrange columns

## Sum BOLD records
boldTree <- boldTree %>% group_by_all() %>% summarise(.groups="keep", counts = n())

## Set non matching species to count 0
boldTree <- within(boldTree, {
  f <- is.na(marker_name) == TRUE
  counts[f] <- NA
  f <- NULL
})

## Add Naturalis records
natTree <- left_join(nsrBackbone[c(1:7)],
                       natCoverage[c('species_name','counts','database_name')],
                       by=c("species"="species_name"))
natTree$marker_name <- NA
colnames(natTree)[7] <- "species_name"

## Merge trees
combTree <- merge(boldTree, natTree,all=TRUE)
combTree <- unique(combTree)
combTree <- combTree %>% drop_na(counts)

## Add backbone to tree
finalTree <- left_join(nsrBackbone[c(1:7)], combTree,
                      by=c("kingdom","phylum","class","order","family","genus",
                           "species"="species_name"))
finalTree <- finalTree[, c(1,2,3,4,5,6,7,10,9,8)]

## Save workspace image
save.image(sprintf('%s/script/workspace.RData', REPO_HOME))
```

### Perform visualisation

Compute and draw reactive shiny filters through collapsible d3js tree

```{r tree-visualisation}
## Define tree data ----
m=finalTree%>%data.frame%>%mutate(NEWCOL=NA)%>%distinct

## Define UI for application ----
ui <- fluidPage(
  fluidRow(
    column(6,
           uiOutput("Hierarchy"),
           tableOutput("clickView"),
           d3treeOutput(outputId="d3", height="800px")
    ),
    column(6,
           "Reactive Table",
           div(DT::dataTableOutput(outputId="table"),
               style = "font-size:80%")
    )
  ),

  hr(),

  fluidRow(
    column(8,billboarderOutput(outputId="billboarder")
    ),
    column(4,billboarderOutput(outputId="markerPie"))
  )
)

## Define server logic ----
server <- function(input, output, session) {
  
  # Tree Hierarchy ----
  output$Hierarchy <- renderUI({
    Hierarchy=names(m)
    Hierarchy=head(Hierarchy,-1)
    selectizeInput("Hierarchy","Tree Hierarchy",
                   choices = Hierarchy,multiple=T,selected = Hierarchy,
                   options=list(plugins=list('drag_drop','remove_button')))
  })
  
  # Observer ----
  network <- reactiveValues()
  
  observeEvent(input$d3_update,{
    network$nodes <- unlist(input$d3_update$.nodesData)
    activeNode<-input$d3_update$.activeNode
    if(!is.null(activeNode)) network$click <- jsonlite::fromJSON(activeNode)
  })
  
  observeEvent(network$click,{
    output$clickView<-renderTable({
      as.data.frame(network$click)
    },caption='Last Clicked Node',caption.placement='top')
  })
  
  
  TreeStruct=eventReactive(network$nodes,{
    df=m
    if(is.null(network$nodes)){
      df=m
    }else{
      
      x.filter=tree.filter(network$nodes,m)
      df=ddply(x.filter,.(ID),function(a.x){m%>%filter_(.dots = list(a.x$FILTER))%>%distinct})
    }
    df
  })
  
  # Render tree
  observeEvent(input$Hierarchy,{
    output$d3 <- renderD3tree({
      if(is.null(input$Hierarchy)){
        p=m
      }else{
        p=m%>%select(one_of(c(input$Hierarchy,"NEWCOL")))%>%unique
      }
      
      d3tree(data = list(root = df2tree(struct = p,rootname = 'm'), layout = 'collapse'),
             activeReturn = c("kingdom", "phylum", "class", "order", "family", "genus",
                              "species","counts", "database_name", "marker_name" ))
    })
  })
  
  observeEvent(network$nodes,{
    output$results <- renderPrint({
      str.out=''
      if(!is.null(network$nodes)) str.out=tree.filter(network$nodes,m)
      return(str.out)
    })    
  })
  
  # Reactive table
  output$table <- DT::renderDataTable(expr = {
    TreeStruct()%>%select(-NEWCOL)
  },
  options=list(scrollX=T,title="Reactive Table")
  )

  # Grahps ----
  output$billboarder <- renderBillboarder(expr = {
    dfActive <- TreeStruct()%>%select(-NEWCOL)
    nsrTotal <- unlist(lapply(dfActive[2:8], function(x) length(unique(x))))
    
    boldActive <- subset(dfActive,database_name=="BOLD")
    boldTotal <- unlist(lapply(boldActive[2:8], function(x) length(unique(x))))
    boldCov <- as.vector(round((boldTotal/nsrTotal * 100),1))
    
    natActive <- subset(dfActive,database_name=="NATURALIS")
    natTotal <- unlist(lapply(natActive[2:8], function(x) length(unique(x))))
    natCov <- as.vector(round((natTotal/nsrTotal * 100),1))
    
    combActive <- subset(dfActive,database_name=="BOLD" | database_name=="NATURALIS")
    combActive <- combActive[!duplicated(combActive$species),]
    combTotal <- unlist(lapply(combActive[2:8], function(x) length(unique(x))))
    combCov <- as.vector(round((combTotal/nsrTotal * 100),1))
    
    vis.cov <- data.frame(
      taxa = paste(names(dfActive)[2:8]),
      bold_total = boldCov,
      nat_total = natCov,
      comb_total = combCov
    )
    
    # Barchart Taxa
    billboarder() %>%
      bb_barchart(
        data = vis.cov[, c("taxa", "bold_total", "nat_total", "comb_total")]
      ) %>%
      bb_data(
        names = list(bold_total = "bold", nat_total = "naturalis", comb_total = "combined")
      ) %>%
      bb_y_grid(show = TRUE) %>%
      bb_y_axis(tick = list(format = suffix("%")),
                label = list(text = "Percentage of taxa covered", position = "outer-top")) %>%
      bb_legend(position = "inset", inset = list(anchor = "top-right")) %>%
      bb_labs(title = "Coverage of taxa")
  })
  
  # Marker Pie Chart
  output$markerPie <- renderBillboarder(expr = {
    dfActive <- TreeStruct()%>%select(-NEWCOL)
    vis.marker <- setNames(aggregate(dfActive$counts, list(dfActive$marker_name),
                                     FUN=sum), c("marker_name", "counts"))
    
    billboarder() %>%
      bb_piechart(data = vis.marker) %>%
      bb_labs(title = "Share of markers (BOLD)")
  })
  
}

shinyApp(ui, server)
```
