---
title: "Custom Databases Analyses"
author: "Bastiaan"
date: "11 January 2021"
output: html_document
---

Introduction
------------

The following R markdown file performs the steps needed to search out and visualize overlap / gaps between obtained public sequence data (using the custom_databases.py script) and its reference by the Dutch Species Register.

The following libraries and scripts need to be loaded:

```{r libraries, include=FALSE}
library(rmarkdown)
library(taxize)
library(myTAI)
library(dplyr)
library(tidyr)
library(collapsibleTree)
library(DT)
```


In addition, We define the global variables that allow us to locate our data:

```{r globals, include=FALSE}
# This should define the root folder of the local copy of the git repository at: https://github.com/naturalis/Custom-databases-DNA-sequences, which is
# automatically defined correctly if we run the present code from within a
# local clone of the repo and have set the working directory to the script
# source (in RStudio: Session > Set working directory > To source file location)

REPO_HOME <- paste(dirname(getwd()), sep = "")
```


Gathering data
--------------

### Reference Data (NSR)

Ideally a full taxonomic hierarchy of species from the NSR would be used as reference. As only the scientific names has been given, classification of these species can be extrated from a database of choice. Records will be parsed into a dataframe.

```{r NSR, include=FALSE}
# Locate NSR species
NSR_Species <- sprintf('%s/data/NSR_exports/NSR_species.txt', REPO_HOME)
nsr_lst <- read.csv(NSR_Species, header=FALSE)

# Extrat taxonomic classification
# Change DB to either ncbi, itis, gbif, bold
nsr_lstoutput <- apply(nsr_lst, 1, function(x) classification(sci_id=x, db="bold", rows=1))

# Parse out the taxonomy levels that you require
nsr_ldata <- unlist(nsr_lstoutput,recursive=FALSE,use.names=TRUE)
df_nsr <- tibble(names = names(nsr_ldata), nsr_ldata) %>% 
  unnest() %>% 
  filter(rank %in% c("phylum","class","order","family","genus","species")) %>% 
  select(-id) %>% 
  spread(rank, name) %>% 
  select(name = names, phylum, class, order, family, genus, species)
df_nsr$name <- NULL


```


### Public Sequence Data (BOLD)

We now load the list of taxa for the obtained public sequence data.

```{r BOLD, include=FALSE}
bold.file <- sprintf('%s/data/FASTA_files/match.fasta', REPO_HOME)

# Taxa to table
df_bold <- read.csv(bold.file, sep = "\t", row.names=NULL)[c("phylum_name","class_name","order_name","family_name","genus_name","species_name")]
```


Computing plots
---------------

### Counting coverage

Calculating the coverage of the public sequence data against its reference.

```{r Coverage, include=FALSE}
coverage <- as.data.frame(table(factor(df_bold$species_name,levels=unique(df_nsr$species))))
colnames(coverage) <- c("species","bold_records")

df3 <- dplyr::left_join(df_nsr, coverage, by = "species")

```


### Draw Tree

```{r Tree, warning=FALSE}
# Compute and draw a collapsible tree
collapsibleTreeSummary(
  df3,
  hierarchy = c("phylum", "class", "order", "family", "genus", "species"),
  height = 1000,
  width = 1500,
  zoomable = TRUE,
  attribute = "bold_records",
  maxPercent = 50
)

datatable(df3, options = list(pageLength = 5))
```