---
title: "Custom Databases: Classification & Analyses"
author: "Bastiaan"
date: "27 March 2021"
output: html_document
runtime: shiny
---

Introduction
------------

The following R markdown file performs the steps needed to:

1. Obtain taxonomic information and Naturalis specimen records for selected taxa,
2. Create a data structure capable of providing highly visual representations
and at the same time maintaining integrity of each data set’s origin,
3. Search out and visualize the overlap / discrepancies between obtained public
reference data and records readily available at Naturalis Biodiversity Center.

The following libraries need to be loaded:

```{r libraries, include=FALSE, message=FALSE, warning=FALSE}
library(rmarkdown)
library(data.table)
library(taxizedb)
library(myTAI)
library(tidyr)
library(shiny)
library(DT)
library(plyr)
library(dplyr)
library(stringr)
library(d3Tree)
library(billboarder)
library(nbaR)
```

In addition, here We define the global variables that we will reuse throughout
the code:

```{r globals, include=FALSE}
# This should define the root folder of the local copy of the git repository at:
# https://github.com/naturalis/Custom-databases-DNA-sequences, which is
# automatically defined correctly if we run the present code from within a
# local clone of the repo and have set the working directory to the script
# source (in RStudio: Session > Set working directory > To source file location)
REPO_HOME <- paste(dirname(getwd()), sep = "")
```

### Load data

We now load the following data sets:

1. Reference Data Set (NSR),
2. Obtained Public specimen records and sequence data (BOLD),
3. List of Marker Codes,
4. List of Public Databases

Both the reference data set (1) and the obtained public reference data (2) are
output from the [script](../script/custom_databases.py)

```{r datasets, include=FALSE}
## Load species and synonyms from the NSR files, respectively
nsrSpecies.file <- sprintf('%s/results/nsr_species.csv', REPO_HOME)
nsrSpecies <- read.csv(nsrSpecies.file, encoding="UTF-8")

nsrSynonyms.file <- sprintf('%s/results/nsr_synonyms.csv', REPO_HOME)
nsrSynonyms <- read.csv(nsrSynonyms.file, encoding="UTF-8")

## Load public reference data
## (Set missing values to na)
bold.file <- sprintf('%s/data/FASTA_files/match.tsv', REPO_HOME)
bold <- fread(bold.file, header=TRUE, sep="\t", na.strings="NA",
               encoding="UTF-8", fill=TRUE, data.table=FALSE)

## Load marker data set
markers.file <- sprintf('%s/results/markers.csv', REPO_HOME)
markers <- read.csv(markers.file, header=TRUE, sep=",")

## Load database data set
databases.file <- sprintf('%s/results/databases.csv', REPO_HOME)
databases <- read.csv(databases.file, header=TRUE, sep=",")
```

### NSR Synonyms

To cover misalignments in identification between databases, known synonyms for
NSR taxa are matched to their corresponding species IDs.

```{r nsrSynonyms, include=FALSE}
## Merge respective taxon data with the synonym data set
## Common values: respective species name and authority with year of publication
nsrSynonyms <- inner_join(nsrSynonyms, nsrSpecies, by=c("taxon_name"="species_name",
           "taxon_author"="identification_reference"))

## Add a unique identifier for each known synonym
## Rearrange and remove obsolete columns
nsrSynonyms <- within(nsrSynonyms, {
  synonym_id <- seq.int(nrow(nsrSynonyms))
}) %>% .[,c(6,5,1,2)]

## Write changes to respective CSV file
write.csv(nsrSynonyms, file=sprintf('%s/results/nsr_synonyms.csv', REPO_HOME),
          row.names=FALSE, fileEncoding="utf-8")
```

### NSR Backbone

To facilitate taxonomic estimation by lowest common ancestor approach, the 
complete classification of species should ideally be present in a reference 
taxonomic data set. As the NSR export only held the scientific names of all
species of interest, the Netherlands Biodiversity API will be queried to obtain
their higher classification.

```{r nsrBackbone, include=FALSE}
## Select species for taxon data retrieval
nsrIn <- tolower(nsrSpecies[, "species_name"])

## Query NSR taxon records (duration: ~40m)
## Method: nbaR 'taxon_query'
## Parameters: NSR source code and list of selected species
queryParams <- list("sourceSystem.code"="NSR")
nsrSpecimens <- data.frame()
for(x in 1:length(nsrIn)){
  queryParams[["acceptedName.scientificNameGroup"]] <- nsrIn[x]
  nsrSpecimens <- bind_rows(nsrSpecimens, taxon_query(queryParams))
}

## Access and extract required taxonomic fields from retrieved taxon records
nsrBackbone = data.frame()
for(x in 1:nrow(nsrSpecimens)){
  tryCatch({
    c.id=nsrSpecimens$sourceSystemId[x]
    c.kingdom=nsrSpecimens$defaultClassification$kingdom[x]
    c.phylum=nsrSpecimens$defaultClassification$phylum[x]
    c.class=nsrSpecimens$defaultClassification$className[x]
    c.order=nsrSpecimens$defaultClassification$order[x]
    c.family=nsrSpecimens$defaultClassification$family[x]
    c.genus=nsrSpecimens$defaultClassification$genus[x]
    c.species=nsrSpecimens$acceptedName$specificEpithet[x]
    c.authority=nsrSpecimens$acceptedName$authorshipVerbatim[x]

    row <- data.frame(cbind(tax_id=c.id,kingdom=c.kingdom,phylum=c.phylum,class=c.class,
                        order=c.order,family=c.family,genus=c.genus,
                        species=paste(c.genus,c.species),identification_reference=c.authority))
    nsrBackbone <- bind_rows(nsrBackbone, row)
  }, error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
}

## Remove parenthesis around the authority for consistency across data sets
nsrBackbone$identification_reference = gsub("[()]", "", nsrBackbone$identification_reference)

## Match taxonomic data to respective species ID, by species name
nsrBackbone <- left_join(nsrBackbone, nsrSpecies[1:2], by=c("species"="species_name")) %>%
  .[,c(1,10,2,3,4,5,6,7,8,9)]
nsrBackbone <- unique(nsrBackbone)

## Write NSR data to file
write.csv(nsrBackbone, file=sprintf('%s/results/tree_nsr.csv', REPO_HOME),
          row.names=FALSE, fileEncoding="utf-8")

## Clean up query variables
rm(x, row, queryParams, list = ls()[grep("^c.", ls())])
```

### Naturalis Coverage

To determine the overlap / discrepancies between the obtained public sequence data
and the reference at Naturalis Biodiversity Center, in relation to the NSR, we now
query the Naturalis specimen records (using the Netherlands Biodiversity API).

```{r natCoverage, include=FALSE}
## Select species for taxon data retrieval
## Ensuring to select only binomial names
natIn <- unique(gsub("([A-Za-z]+).*", "\\1", nsrSpecies$species_name))

## Query Naturalis specimen records (duration: ~20m)
## Method: nbaR 'specimen_query'
## Parameters: CRS source code and list of selected species
queryParams <- list(queryParams <- list("sourceSystem.code"="CRS"))
natSpecimens <- data.frame()
for(x in 1:length(natIn)){
  queryParams[["identifications.defaultClassification.genus"]] <- natIn[x]
  natSpecimens <- bind_rows(natSpecimens, specimen_query(queryParams))
}

## Access and extract required data fields from retrieved specimen records
natSpecies = data.frame()
for(x in 1:nrow(natSpecimens)){
  tryCatch({
    c.genus=natSpecimens$identifications[[x]]$defaultClassification$genus
    c.species=natSpecimens$identifications[[x]]$defaultClassification$specificEpithet
    c.count=natSpecimens$numberOfSpecimen[[x]]
    c.id=natSpecimens$sourceSystemId[[x]]

    row <- data.frame(cbind(species=paste(c.genus,c.species),counts=c.count,sequenceID=c.id))
    natSpecies <- bind_rows(natSpecies, row)
  }, error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
}

## Remove NAs
natSpecies$counts[is.na(natSpecies$counts)] <- 0

## Update data type and sum unique species record counts
natSpecies$counts <- as.numeric(as.character(natSpecies$counts))
natSpecies <- aggregate(natSpecies$counts,
                        by=list(natSpecies$species,natSpecies$sequenceID),sum)

## Rename columns
colnames(natSpecies) <- c("species", "sequenceID", "counts")

## Adopt respective taxonomic unit for synonymous names
natSpecies <- left_join(natSpecies, nsrSynonyms[2:3], by=c("species"="synonym_name"))
natSpecies <- left_join(natSpecies, nsrSpecies[1:2], by=c("species_id"))
natSpecies$species_name <- coalesce(natSpecies$species_name, natSpecies$species)
natSpecies <- natSpecies[,c(5,2,3)]

## Match species against the NSR
natCoverage <- merge(natSpecies, nsrSpecies[c(2:3)], by="species_name")

## Clean up query variables
rm(x, row, queryParams, list = ls()[grep("^c.", ls())])
```

### Additional taxonomy

The ability to divert to a different backbone provides additional insight into a
specimens classification or serves as a filter for specific taxa (e.g., marine: WoRMS).
Additional taxonomic hierarchies can be retrieved from various databases.

```{r tree_ncbi, include=FALSE}
## Download taxonomic database (options: ncbi, itis, gbif, col, wfo)
## Used data source: NCBI
db_download_ncbi()

## Load species names from the NSR data set
taxidIn <- nsrSpecies[, "species_name"]

## Match species names to NCBI taxon IDs
taxidOut <- taxizedb::name2taxid(taxidIn, db="ncbi", out_type="summary")

## Isolate taxon IDs from output, save as vector
treeIn <- as.vector(taxidOut$id)

## Retrieve taxonomic hierarchy for each taxon ID
treeOut <- taxizedb::classification(treeIn, db="ncbi", row=1, verbose=FALSE)

## Parse out the taxonomy levels/factors that you require
taxdata = data.frame()
for(x in 1:length(treeOut)){
  tryCatch({
    for(i in 1:length(treeOut[[x]][[1]])) {
      c.tax_id=filter(treeOut[[x]])$id[[i]]
      c.parent_tax_id=NULL
      c.rank=filter(treeOut[[x]])$rank[[i]]
      c.name=filter(treeOut[[x]])$name[[i]]

      tryCatch({
        c.parent_tax_id=filter(treeOut[[x]])$id[[i-1]]
        }, error=function(e){c.parent_tax_id=NULL})
      
      row <- data.frame(cbind(tax_id=c.tax_id,parent_tax_id=c.parent_tax_id,
                            rank=c.rank, name=c.name))

      # Check if record exists
      if (nrow(merge(row,taxdata))==0) {
        taxdata <- bind_rows(taxdata, row)
      }
    }
  }, error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
}

## Match hierarchies to NSR species names by taxon IDs
## (Accounting for use of synonymous names within NCBI)
ncbiBackbone <- left_join(taxdata, taxidOut, by=c("tax_id"="id"),
                       suffix=c("","_nsr")) %>%
  left_join(., nsrSpecies[1:2], by=c("name_nsr" = "species_name")) %>%
  .[, c(1,6,4,2,3)] # Reaarange columns, drop 'name_nsr' column
ncbiBackbone <- unique(ncbiBackbone)

## Write NCBI data to file
write.csv(ncbiBackbone, file=sprintf('%s/results/tree_ncbi.csv', REPO_HOME),
          row.names=FALSE, fileEncoding="utf-8", na="")

## Clean up query variables
rm(x, i, row, list = ls()[grep("^c.", ls())])
```

### Create data structure

Following NSR’s species names (and known synonyms) for the identification of all
molecular data, the underlying data structure combines the molecular data through
NSR’s accepted species names and links sequence information to its taxonomic data.

```{r species_markers, include=FALSE}
## Match the NSR species names with the obtained (BOLD) public reference data
## Subsequently, fetch the associated ID for each markercode
species_markers <- right_join(nsrSpecies, bold[c("species_name","markercode","sequenceID",
                                                 "identification_reference")],
                              by=c("species_name", "identification_reference")) %>%
  right_join(., markers, by=c("markercode" = "marker_name"))

## Asign database ID and update data type of the sequenceID
species_markers$database_id <- 1
species_markers$sequenceID <- as.character(species_markers$sequenceID)

## Prepare merge with Naturalis records by aligning data
naturalis <- left_join(natCoverage, nsrSpecies, by=c("species_name",
                                                     "identification_reference"))
naturalis$database_id <- 0
naturalis$marker_id <- 0
naturalis$counts <- NULL

## Merge BOLD and Naturalis records
species_markers <- merge(species_markers, naturalis,
                         by=c("species_id","identification_reference","sequenceID",
                              "database_id","marker_id","species_name"), all=TRUE)

## Add a unique identifier for each record
species_markers <- within(species_markers, {
  species_marker_id <- seq.int(nrow(species_markers))
}) %>%
  .[, c(8,1,4,5,3)] %>% # Remove obsolete and rearrange columns
  .[complete.cases(.[ , 2]),] # Omit empty species_id's (NAs)

## Write data as csv to output folder
write.csv(species_markers, file=sprintf('%s/results/species_markers.csv', REPO_HOME),
          row.names=FALSE, fileEncoding="utf-8")
```

### Load Tree

To perform an analyses on our species we now load the backbone of the NSR. This 
is their complete tree, populated with our collected data. Here where we seek
out overlap / gaps between Naturalis and our obtained Public Sequence Data.

```{r tree, include=FALSE}
## Get classification for each obtained record
## Subsequently, fetch the associated marker and database names for each record
tree <- left_join(nsrSpecies[1], nsrBackbone[c(2:9)], by=c("species_id")) %>%
  left_join(., species_markers[c("species_id","database_id", "marker_id")],
            by=c("species_id")) %>%
  left_join(., markers, by=c("marker_id")) %>%
  left_join(., databases, by=c("database_id"))

## Sum count of unique records
tree <- tree %>% group_by_all() %>% summarise(.groups="keep", counts = n())

## Set count for non matching species to NA
tree <- within(tree, {
  f <- is.na(database_name) == TRUE
  counts[f] <- NA
  f <- NULL
})

## Remove obsolete and rearrange columns
tree <- tree[,c(2,3,4,5,6,7,8,11,12,13)]

## Save workspace image
save.image(sprintf('%s/.RData', REPO_HOME))
```

### Perform visualisation

A dynamic visualization of this complex data set is presented by a collapsible
Reingold-Tilford tree diagram. Through Shiny, the server observes the d3
collapsible tree library and its real-time layout. Data transferred back to Shiny
is mapped to a series of logical expressions to create reactive filters.
Various charts visualize the number of occurrence records and its relevant data.

```{r tree-visualisation}
## Define tree data ----
m=tree%>%data.frame%>%mutate(NEWCOL=NA)%>%distinct

## Define UI for application ----
ui <- fluidPage(
  fluidRow(
    column(7,
           uiOutput("Hierarchy"),
           tableOutput("clickView"),
           d3treeOutput(outputId="d3", height="800px")
    ),
    column(5,
           "Reactive Table",
           div(DT::dataTableOutput(outputId="table"),
               style = "font-size:80%")
    )
  ),

  hr(),

  fluidRow(
    column(12,
           "Records per taxa",
           div(DT::dataTableOutput(outputId="coverage"),
               style = "font-size:80%")
    )
  ),

  hr(),

  fluidRow(
    column(8,billboarderOutput(outputId="billboarder")
    ),
    column(4,billboarderOutput(outputId="markerPie"))
  )
)

## Define server logic ----
server <- function(input, output, session) {

  # Tree Hierarchy ----
  # Enable hierarchy order of the levels to be switched through its input box objects
  output$Hierarchy <- renderUI({
    Hierarchy=names(m)
    Hierarchy=head(Hierarchy,-1)
    selectizeInput("Hierarchy","Tree Hierarchy",
                   choices = Hierarchy,multiple=T,selected = Hierarchy,
                   options=list(plugins=list('drag_drop','remove_button')))
  })

  # Observer ----
  # The 'activeNode' observer returns meta data of the last clicked node
  network <- reactiveValues()

  observeEvent(input$d3_update,{
    network$nodes <- unlist(input$d3_update$.nodesData)
    activeNode<-input$d3_update$.activeNode
    if(!is.null(activeNode)) network$click <- jsonlite::fromJSON(activeNode)
  })

  observeEvent(network$click,{
    output$clickView<-renderTable({
      as.data.frame(network$click)
    },caption='Last Clicked Node',caption.placement='top')
  })

  # If a node is clicked it is interpreted as being of interest and a logical
  # expression is created to return it and its children.
  # If specific siblings are clicked, and opened, then the non-clicked siblings are not
  # returned.
  TreeStruct=eventReactive(network$nodes,{
    df=m
    if(is.null(network$nodes)){
      df=m
    }else{

      x.filter=tree.filter(network$nodes,m)
      df=ddply(x.filter,.(ID),function(a.x){m%>%filter_(.dots = list(a.x$FILTER))%>%distinct})
    }
    df
  })

  # Render tree ----
  # Meta data from the observer is inputed in the d3tree call in the activeReturn argument
  observeEvent(input$Hierarchy,{
    output$d3 <- renderD3tree({
      if(is.null(input$Hierarchy)){
        p=m
      }else{
        p=m%>%select(one_of(c(input$Hierarchy,"NEWCOL")))%>%unique
      }

      d3tree(data = list(root = df2tree(struct = p,rootname = 'm'), layout = 'collapse'),
             activeReturn = c("kingdom", "phylum", "class", "order", "family", "genus",
                              "species","counts", "database_name", "marker_name" ))
    })
  })

  # Filters mapped from tree ----
  observeEvent(network$nodes,{
    output$results <- renderPrint({
      str.out=''
      if(!is.null(network$nodes)) str.out=tree.filter(network$nodes,m)
      return(str.out)
    })
  })

  # Reactive table ----
  # Creates a filtered datatable, containing the selected nodes
  output$table <- DT::renderDataTable(expr = {
    TreeStruct()%>%select(-NEWCOL)
  },
  options=list(scrollX=T)
  )
  
  # Calculating coverage of selected nodes
  calcCoverage <- function() {
    dfActive <- TreeStruct()%>%select(-NEWCOL)
    nsrTotal <- unlist(lapply(dfActive[2:8], function(x) length(unique(x))))
    
    boldActive <- subset(dfActive,database_name=="BOLD")
    boldTotal <- unlist(lapply(boldActive[2:8], function(x) length(unique(x))))
    boldCov <- as.vector(round((boldTotal/nsrTotal * 100),1))
    
    natActive <- subset(dfActive,database_name=="NATURALIS")
    natTotal <- unlist(lapply(natActive[2:8], function(x) length(unique(x))))
    natCov <- as.vector(round((natTotal/nsrTotal * 100),1))
    
    combActive <- subset(dfActive,database_name=="BOLD" | database_name=="NATURALIS")
    combActive <- combActive[!duplicated(combActive$species),]
    combTotal <- unlist(lapply(combActive[2:8], function(x) length(unique(x))))
    combCov <- as.vector(round((combTotal/nsrTotal * 100),1))

    coverageDF <- data.frame(
      taxa = paste(names(dfActive)[2:8]),
      bold_total = boldTotal,
      nat_total = natTotal,
      comb_total = combTotal,
      bold_cov = boldCov,
      nat_cov = natCov,
      comb_cov = combCov
    )
    return(coverageDF)
  }
  
  # Creates coverage table
  output$coverage <- DT::renderDataTable(expr = {
    calcCoverage() %>% select(2:4)
  },
  options=list(scrollx=F, paging=F, lengthChange=F,
               searching=F,  title="Records per taxa"),
  filter=c("none"),
  autoHideNavigation=FALSE
  )

  # Grahps ----
  # Plot barchart for taxon coverage
  output$billboarder <- renderBillboarder(expr = {
    coverageDF <- calcCoverage()
    billboarder() %>%
      bb_barchart(
        data = coverageDF[, c("taxa", "bold_cov", "nat_cov", "comb_cov")]
      ) %>%
      bb_data(
        names = list(bold_cov = "bold", nat_cov = "naturalis", comb_cov = "combined")
      ) %>%
      bb_y_grid(show = TRUE) %>%
      bb_y_axis(tick = list(format = suffix("%")),
                label = list(text = "Percentage of taxa covered", position = "outer-top")) %>%
      bb_legend(position = "inset", inset = list(anchor = "top-right")) %>%
      bb_labs(title = "Coverage of taxa")
  })

  # Plot pie chart for share of markers
  output$markerPie <- renderBillboarder(expr = {
    dfActive <- TreeStruct()%>%select(-NEWCOL)
    markerDF <- setNames(aggregate(dfActive$counts, list(dfActive$marker_name),
                                   FUN=sum), c("marker_name", "counts"))
    
    billboarder() %>%
      bb_piechart(data = markerDF) %>%
      bb_labs(title = "Share of markers (BOLD)")
  })

}

shinyApp(ui, server)
```
